### 2018-07-18 source/Ch0-Foreword/foreword.tex
final commit for sake of backup - thesis completed

\pg
What are radio waves? Electromagnetic waves were theorised by James Clerk Maxwell \citep{maxwell} in his great theoretical contribution to modern physics, their speed matching the speed of light as measured by Ole Christensen R\o mer and, later, James Bradley. It was not until Heinrich Rudolf Hertz's 1887 experiment that these waves were measured in a laboratory, leading to the dawn of radio communications - and, later, radio astronomy. The link between radio waves and light was one of association: light is known to behave as a wave (Young double-slit experiment), with the same propagation speed as electromagnetic radiation. Light ``proper" is also known to exist beyond the optical regime: Herschel's experiment shows that when diffracted through a prism, sunlight warms even those parts of a desk which are not observed to be lit (first evidence of infrared light). The link between optical light and unseen electromagnetic radiation is then an easy step, and one confirmed through countless technological applications (e.g. optical fiber to name but one). And as soon as this link is established, a question immediately comes to the mind of the astronomer: what does the sky, our Universe, look like to the radio ``eye"?

>>>>

\pg
What are radio waves? Electromagnetic waves were theorised by James Clerk Maxwell \citep{maxwell} in his great theoretical contribution to modern physics, their speed matching the speed of light as measured by Ole Christensen R\o mer and, later, James Bradley. It was not until Heinrich Rudolf Hertz's 1887 experiment that these waves were measured in a laboratory, leading to the dawn of radio communications - and, later, radio astronomy. The link between radio waves and light was one of association: light is known to behave as a wave (Young double-slit experiment), with the same propagation speed as electromagnetic radiation. Light ``proper" is also known to exist beyond the optical regime: Herschel's experiment shows that when diffracted through a prism, sunlight warms even those parts of a desk which are not observed to be lit (first evidence of infrared light). The link between optical light and unseen electromagnetic radiation is then an easy step to make, and one confirmed through countless technological applications (e.g. optical fiber to name but one). And as soon as this link is established, a question immediately comes to the mind of the astronomer: what does the sky, our Universe, look like to the radio ``eye"?

### 2018-07-18 source/Ch1-ThisThesis/intro.tex
final commit for sake of backup - thesis completed

\subsection{Active Galactic Nuclei}

\pg
The underlying motivation of all the work presented in this manuscript is the study of radio-active galactic nuclei, or radio AGNs. A radio AGN is a type of radio galaxy whose emission is dominated by its nucleus. This nucleus emission is non-thermal, and assumed to be emitted from both collimated jets springing from the accretion of matter onto a galaxy's supermassive black hole and from a hot accretion disk surrounding said black hole. AGNs are then commonly split into different types, based on spectral line behaviour \citepads[as in]{1943ApJ....97...28S} or morphological characteristics \citepads[as in]{1974MNRAS.167P..31F}.

\pg
So-called unified AGN models, which aim to explain the underlying physical object which is assumed to lie behind all observed examples of AGN, were developed to explain the diversity in observed AGN types \citepads[e.g.]{1985ApJ...297..621A,2008A&A...490..893T,2015ARA&A..53..365N,2017MNRAS.469.1883M}. These models would include a central engine, usually a supermassive black hole with hot accretion \citepads[innermost accretion regions estimated to be around $10^7$K by]{1972A&A....21....1P}, emitting thermally; an optically disk torus, typically of size 10 to 100pc, blocking light emitted from the central engine along the equitorial plane, and the cavity inside which would be the region from which broad emission lines would emit; and finally, ionisation cones in the direction of the poles, where the emission from the central engine would collimate into large-scale jets (which would also, at the scale of a few tens of parsecs, create the region of narrow spectral line emission).

\pg
The interaction between AGN jets and the surrounding intergalactic medium can result in structure such as that shown in \cref{fig.agn} below: a radio galaxy in the centre of two extremely radio-bright jets, with lots of diffuse emission from their lobes. These lobes show where the accretion-accelerated particles begin to slow down - and therefore emit - due to contact with the intergalactic medium. Hence, their sometimes peculiar shape: this medium can be complex and inhomogeneous (if the radio galaxies lie in a cluster, for example) which can lead to the sort of jet structure shown in this image.
\begin{figure}[!h]
\centering
\includegraphics[width=0.6\textwidth]{images/{3C31_radio_optical_montage_Alan_Bridle}.png}
\caption{\label{fig.agn} Radio optical montage of 3C31, by Alan Bridle.}
\end{figure}

\pg
Of course, not all galaxies have active nuclei, and those which do also have other forms of radio emission: for example, free-free radiation from the interstellar medium, ionised by light from newborn stars, will often contribute to radio-galactic flux. Some galaxies with active nuclei do not have strong jet emission: these are called radio-quiet AGNs. We can thus identify two components to a radio AGN: the host galaxy and its jets. Peak star formation is thought to have occurred around $z\sim2$, and so the study of AGN activity around this period is of strong scientific interest - could the jets from active AGNs have contributed to the decline in star formation? Were there large populations of older AGNs? These are the astrophysical questions which motivate the study of AGN sources \citepads{2007NewAR..51..168B,2018arXiv180501544X,2018ApJ...853...24U}, which is the driving scientific concern which motivates the present doctorate.
%
%\pg
%Both of these components tell us different things about the galaxy and its surrounding medium. If the host galaxy can be resolved out from the jets, for example, it becomes possible to study the star formation history of that galaxy by comparing the radio flux at different frequencies to the galaxy's brightness in other bands (X, IR, optical etc). Radio waves are not subject to extinction in the way that infrared or optical emission can be: this fact needs to be taken into account when performing this sort of analysis.
%
%\pg

@@

\pg
The work done in this thesis can be summarised as the development of a new algorithm for mitigating calibration artefacts in radio-interferometric images, and its application to a peculiar famous extragalactic field. Because our data volumes put radio interferometry in the domain of ``Big Data" (there are $\sim$245TB of data on the EGS alone), there is a very real need for fast, reliable methods with which to improve interferometric images. This algorithm follows from an improved understanding of the noise properties of interferometric images, and can be understood as a way to better constrain the flux distribution in images by minimising the presence of artefacts. This, in turn, helps self-calibration convergence by improving the image resulting from each imaging step of the self-calibration process.

@@

\pg
Making this image is only possible because LOFAR observations of the Extended Groth Strip includes an extremely bright source, 3C295, in their field of view. This is both a blessing and a curse: unless this source is properly modeled and subtracted, it will pollute images of the EGS. However, because it is so bright, it can serve as a good calibrator source for the LOFAR international stations - if a good enough model of 3C295 can be made and used to calibrate these stations, then high-resolution images of the EGS becomes attainable. This would give access to a large population of radio AGNs at high resolution, observed at a large range of frequencies, and many of which lie at redshifts near the peak of star formation; this improvement of our low-frequency understanding of these objects could, in turn, lead to an improved understanding of these objects more generally.

@@

\pg
A significant portion of the work done as part of this thesis was on investigating the relationship between calibration gain errors and calibration artefacts in the image-plane. What started as an attempt to properly write a weighting scheme to down-weigh poorly-calibrated visibilities eventually became a formal relationship between these quantities, showing both the behaviour of image-plane variance as a function of direction but also of pixel covariance over the full image. This work is shown in much greater detail in \cref{chapter.paper}.

\pg
The key finding of our algorithmic work is a fundamental relationship between the covariance matrices of residual visibilities and the map of the covariance in the image-plane: the ``Cov-Cov relationship'' between visibility covariance to image-plane covariance. The pixel variance and covariance in the image-plane are determined by a ``noise-PSF'', convolved with every pixel in the sky. This also means that thermal noise in interferometric images is correlated between pixels, and that this correlation is given exactly by the PSF: in the limit of an empty, non-deconvolved sky, we therefore properly model the thermal noise properties of its interferometric images. This noise-PSF is the product of the Fourier transforms of the calibration gain covariance matrix with each cell mapped not from $uv$ space to $lm$ coordinates but rather between their respective covariance spaces - from a new {differential Fourier plane (henceforth ``$\left(\deltu\deltv\right)$''-plane) to the image-plane covariance space $\deltl\deltm$. This image-plane covariance space describes} the variance in each pixel and the covariance between pixels\footnote{{The noise-PSF also relates $\delta w$ to $\delta n$, as shown in the matrix formalism, but this is not explicitly referenced in the text since visibility space is usually referred to as ``the UV-plane'' in literature, rather than ``the UVW-space''.}}. It describes the expected calibration artefacts and thermal noise around each source, does not vary as a function of direction, and is convolved with each source in the field to yield the final error map. Because all unwanted (in our case, unphysical) signal can be thought of as noise, we will refer to {the pixel variance map} as the ``noise-map''.

@@

\pg
In this section, we briefly describe previous observations of the Extended Groth Strip (EGS), an extragalactic field with a rich multi-wavelength coverage described in Table \ref{table.EGS.observation}. It has been long observed as part of the All-Wavelength Extended Groth Strip International Survey collaboration (AEGIS, \citetads{2007ApJ...660L...1D}), which later became part of the CANDELS collaboration (\citetads{2011ApJS..197...35G}). The field is centred at $\alpha=14^h17^m,\delta=+52\deg 30'$, placing it between the tail of Ursa Major and Draco. Its size is $0.7'\times 0.1'$. It has notably been the subject of very deep Hubble Space Telescope observations, which the international LOFAR could attempt to match.

@@

\pg
This work, and its results, are shown in \cref{section.EGS.lowres}. Note that the AEGIS observations summarised in \cref{table.EGS.observation} do not cover the full LOFAR primary beam. The sources observed at low resolution are therefore matched with Sloan Digital Sky Survey (SDSS, \citetads{2000AJ....120.1579Y}) images. Overlays are shown to highlight the wealth of interesting sources, and their optical counterparts, which can be picked up as a byproduct of producing high-resolution maps of the radio sky. We select 12 particularly interesting sources and show the results of source matching for these candidates. Overlays of the same SDSS images with NVSS (NRAO VLA Sky Survey, \citetads{1998AJ....115.1693C}) postage stamps are also shown for comparison. Many of the chosen sources do not lie within the Extended Groth Strip, and are therefore only imaged at low resolution, but follow-up images and observations could be made should the need arise.

@@

\pg
Before launching the EGS imaging run using all calibrated datasets, we measure the impact of direction-dependent effects on known sources in the field. Two direction-dependent effects are expected to dominate: decorrelation and differential gains. The first is modeled by the imager used, whereas the second is a function of the gain variability as a function of distance from the calibrator. Seven calibrator sources, lying at various distances and directions from the calibrator source, are examined. They are summarised in \cref{table.LOBOS.sources}, which is reproduced below.

\begin{table}[h!]
\begin{tabular}{ccccc}
\# & RA [hms]    & Dec [dms]   & Dist. from EGS [deg.] & Dist. from 3C295 [deg.] \\\hline
1  & 14:30:18.72 & 52:17:29.80 & 2.041                         & 2.904 \\
2  & 14:19:44.44 & 54:23:04.58 & 1.928                         & 2.517 \\ 
3  & 14:21:20.05 & 53:03:46.00 & 0.864                         & 1.743 \\
4  & 14:21:09.41 & 51:22:32.46 & 1.294                         & 1.728 \\
5  & 14:11:50.32 & 52:49:02.66 & 0.844                         & 0.619 \\
6  & 14:11:20.23 & 52:12:04.30 & 0.915                         & 0.000 \\
7  & 14:08:07.00 & 52:55:11.36 & 1.409                         & 0.869 \\
8  & 14:08:09.76 & 52:44:46.56 & 1.354                         & 0.680 \\
\end{tabular}
\caption{\label{table.LOBOS.sources1}Table recapitulating the positions of all 8 chosen calibrator sources, along with their distance from the observation phase centre (EGS phase centre) and calibrator phase centre (3C295), respectively.}
\end{table}
\begin{figure}[h!]
\includegraphics[width=\linewidth]{images/{image_full_ampphase_di_m.NS_Smooth.noise01.fitslbcs.egs.positions}.png}
\caption{Position of the calibrator sources around the EGS, overlaid on the widefield EGS image. LBCS is 3C295.}
\label{lbcs-coverage-image1}
\end{figure}
%\begin{figure}[h!]
%\includegraphics[width=0.8\linewidth]{images/EGS_LOBOS_scatterplot}
%\caption{Position of the calibrator sources around the EGS. Note that this is not projected properly onto the celestial sphere.}
%\label{bootes-coverage-image}
%\end{figure}

\pg
If directional gains are measured to have little or no impact, then direction-dependent calibration of the international stations will not be required. This is the ideal scenario, as it does not require the use of fringe fitting \citepads{1984ARA&A..22...97P}. If they are measured to have a significant impact, then direction-dependent calibration will be required for international LOFAR. We expect the angular scale of differential gain evolution to be of the order of a degree, and so most of the EGS should be relatively unaffected.

\pg
Having quantified the impact of directional gain errors on our field, we then proceed to a patchwise imaging of the EGS.

>>>>

\pg
Radio astronomy is a young but extremely dynamic field with a large range of science cases. From the study of the epoch of reionisation, which involves looking for polarised, redshifted 21cm line emission to study the emergence of the first luminous sources\footnote{See \url{https://www.haystack.mit.edu/ast/science/epoch/} for more information - the ultimate goal of this topic is to  learn more about the process of structure formation in the universe: what exactly are the evolutionary links between the very smooth matter distribution revealed by CMB studies and the highly structured universe of galaxies and clusters thereof which we know today?}, to that of transient radio sources such as pulsars and exoplanets, going through the study of ultra-high-energy cosmic rays, solar science, space weather science, the study of cosmic magnetism, and the creation of deep extragalactic surveys, radio-astronomy opens up an extremely fertile window into our Universe.

\pg
In \cref{fig.meerkat.galcentre}, for example, much structure is immediately visible: supernova remnants in the galactic plane (the ``bubbles"), gas following galactic magnetic field lines (the filamentary structure), and emission from diffuse gas structure around the core of the Milky Way. Other sources are also visible, however: some unresolved sources, and very many AGN sources. This is one particularly striking example of what radio telescopes allow us to see of astrophysical sources, but is only one among many.
\begin{figure}[!h]
	\centering
	\includegraphics[width=\textwidth]{images/{MeerKAT-Galactic-Centre-Low-Res}.jpg}
	\caption{\label{fig.meerkat.galcentre} MeerKAT image of the galactic centre, made by SARAO. This is a low-resolution version of the true image.}
\end{figure}

\pg
The work presented here falls under the auspices of the deep extragalactic surveys science project. The underlying scientific motivation behind this Key Science Project or KSP is the study of, among other things, radio-active galactic nuclei, or radio AGNs. A radio AGN is a type of radio galaxy whose emission is dominated by its nucleus. This nucleus emission is non-thermal, and assumed to be emitted from both collimated jets springing from the accretion of matter onto a galaxy's supermassive black hole and from a hot accretion disk surrounding said black hole. AGNs are then commonly split into different types, based on spectral line behaviour \citepads[as in]{1943ApJ....97...28S} or morphological characteristics \citepads[as in]{1974MNRAS.167P..31F}.

\pg
So-called unified AGN models, which aim to explain the underlying physical object which is assumed to lie behind all observed examples of AGN, were developed to explain the diversity in observed AGN types \citepads[e.g.]{1985ApJ...297..621A,2008A&A...490..893T,2015ARA&A..53..365N,2017MNRAS.469.1883M}. These models would include a central engine, usually a supermassive black hole with hot accretion \citepads[innermost accretion regions estimated to be around $10^7$K by]{1972A&A....21....1P}, emitting thermally; an optically thick disk torus, typically of size 10 to 100pc, blocking light emitted from the central engine along the equitorial plane, and the cavity inside which would be the region from which broad emission lines would emit; and finally, ionisation cones in the direction of the poles, where the emission from the central engine would collimate into large-scale jets (which would also, at the scale of a few tens of parsecs, create the region of narrow spectral line emission).

\pg
The interaction between AGN jets and the surrounding intergalactic medium can result in structure such as that shown in \cref{fig.agn} below: a radio galaxy in the centre of two extremely radio-bright jets, with lots of diffuse emission from their lobes. These lobes show where the accretion-accelerated particles begin to slow down - and therefore emit - due to contact with the intergalactic medium. Hence, their sometimes peculiar shape: this medium can be complex and inhomogeneous (if the radio galaxies lie in a cluster, for example) which can lead to the sort of jet structure shown in this image.
\begin{figure}[!h]
\centering
\includegraphics[width=0.6\textwidth]{images/{3C31_radio_optical_montage_Alan_Bridle}.png}
\caption{\label{fig.agn} Radio optical montage of 3C31, by Alan Bridle: \url{https://www.cv.nrao.edu/~abridle/3c31xopt.htm}.}
\end{figure}

\pg
Of course, not all galaxies have active nuclei, and those which do also have other forms of radio emission: for example, free-free radiation from the interstellar medium, ionised by light from newborn stars, will often contribute to radio-galactic flux. Some galaxies with active nuclei do not have strong jet emission: these are called radio-quiet AGNs. We can thus identify two components to a radio AGN: the host galaxy and its jets. Peak star formation is thought to have occurred around $z\sim2$, and so the study of AGN activity around this period is of strong scientific interest. Could the jets from active AGNs have contributed to the decline in star formation? Are there large populations of older AGNs, and if so, how old? These are the astrophysical questions which motivate the study of AGN sources \citepads{2007NewAR..51..168B,2018arXiv180501544X,2018ApJ...853...24U}, which is the driving scientific concern which motivates the present doctorate.
%
%\pg
%Both of these components tell us different things about the galaxy and its surrounding medium. If the host galaxy can be resolved out from the jets, for example, it becomes possible to study the star formation history of that galaxy by comparing the radio flux at different frequencies to the galaxy's brightness in other bands (X, IR, optical etc). Radio waves are not subject to extinction in the way that infrared or optical emission can be: this fact needs to be taken into account when performing this sort of analysis.
%
%\pg

@@

\pg
The work done in this thesis can be summarised as the development of a new algorithm for mitigating calibration artefacts in radio-interferometric images, and its application to a peculiar famous extragalactic field. Because our data volumes put radio interferometry in the domain of ``Big Data" (there are $\sim$245TB of LOFAR data centred on the EGS, for example), there is a very real need for fast, reliable methods with which to improve interferometric images. This algorithm follows from an improved understanding of the noise properties of interferometric images, and can be understood as a way to better constrain the flux distribution in images by minimising the presence of artefacts. This, in turn, helps self-calibration convergence by improving the image resulting from each imaging step of the self-calibration process.

@@

\pg
Making this image is possible only because LOFAR observations of the Extended Groth Strip includes an extremely bright source, 3C295, in their field of view. This is both a blessing and a curse: unless this source is properly modeled and subtracted, it will pollute images of the EGS. However, because it is so bright, it can serve as a good calibrator source for the LOFAR international stations - if a good enough model of 3C295 can be made and used to calibrate these stations, then high-resolution images of the EGS becomes attainable. This would give access to a large population of radio AGNs at high resolution, observed at a large range of frequencies, and many of which lie at redshifts near the peak of star formation; this improvement of our low-frequency understanding of these objects could, in turn, lead to an improved understanding of these objects more generally.

\pg
To really be able to perform wide-field images using the full LOFAR array, we cannot rely on the happy circumstances we find ourselves in for our applications project: a much more thorough and reliable approach is necessary. The work of the LOFAR long-baseline team \citepads[e.g.]{2016A&A...595A..86J} shows that kind of approach: in our project, we test specific strategies and approaches which are complimentary to their work, but cannot be applied more generally without their results.

@@

\pg
A significant portion of the work done as part of this thesis was on investigating the relationship between calibration gain errors and calibration artefacts in the image-plane. What started as an attempt to properly write a weighting scheme to down-weigh poorly-calibrated visibilities\footnote{i.e. to justify weighting visibilities by a function inversely proportional to the variance of the associated residual visibilities.} eventually became a formal relationship between these quantities, showing both the behaviour of image-plane variance as a function of direction but also of pixel covariance over the full image. This work is shown in much greater detail in \cref{chapter.paper}.

\pg
The key finding of our algorithmic work is a fundamental relationship between the covariance matrices of residual visibilities and the covariance of pixels in the image-plane: the ``Cov-Cov relationship'' between visibility covariance and image-plane covariance. The pixel variance and covariance in the image-plane are determined by a ``noise-PSF'', convolved with every pixel in the sky. This also means that thermal noise in interferometric images is correlated between pixels, and that this correlation is given exactly by the PSF: in the limit of an empty, non-deconvolved sky, we therefore properly model the thermal noise properties of its interferometric images. This noise-PSF is the product of the Fourier transforms of the calibration gain covariance matrix with each cell mapped not from $uv$ space to $lm$ coordinates but rather between their respective covariance spaces - from a new {differential Fourier plane (henceforth ``$\left(\deltu\deltv\right)$''-plane) to the image-plane covariance space $\deltl\deltm$. This image-plane covariance space describes} the variance in each pixel and the covariance between pixels\footnote{{The noise-PSF also relates $\delta w$ to $\delta n$, as shown in our matrix formalism, but we do not explicitly reference this in the text for the sake of brevity, as visibility space is usually referred to as ``the UV-plane'' in literature, rather than ``the UVW-space''.}}. It describes the expected calibration artefacts and noise level around each source, does not vary as a function of direction, and is convolved with each source in the field to yield the final error map. Because all unwanted (in our case, unphysical) signal can be thought of as noise, we will refer to {the pixel variance map} as the ``noise-map''.

\pg
Based on this theoretical finding, we develop a weighting scheme algorithm with which to improve the noise properties of interferometric images by down-weighting either visibilities with high variance in their calibration residuals, or visibilities which are strongly correlated to others in their calibration residuals. We implement a robust version of the first weighting scheme, which relies on an antenna-based estimate of the visibility variance.

@@

\pg
In this section, we briefly describe previous observations of the Extended Groth Strip (EGS), an extragalactic field with a rich multi-wavelength coverage described in Table \ref{table.EGS.observation}. It has been long observed as part of the All-Wavelength Extended Groth Strip International Survey collaboration (AEGIS, \citetads{2007ApJ...660L...1D}), which later became part of the CANDELS collaboration (\citetads{2011ApJS..197...35G}). The field is centred at $\alpha=14^h17^m,\delta=+52^o 30'$, placing it between the tail of Ursa Major and Draco. Its size is $0.7'\times 0.1'$. It has notably been the subject of very deep Hubble Space Telescope observations, which the international LOFAR could attempt to match.

@@

\pg
This work, and its results, are shown in \cref{section.EGS.lowres}. Note that the AEGIS observations summarised in \cref{table.EGS.observation} do not cover the full LOFAR primary beam. The sources observed at low resolution are therefore matched with Sloan Digital Sky Survey (SDSS, \citetads{2000AJ....120.1579Y}) images. Overlays are shown to highlight the wealth of interesting sources, and their optical counterparts, which can be picked up as a byproduct of high-resolution maps of the radio sky. We select 12 particularly interesting sources and show the results of source matching for these candidates. Overlays of the same SDSS images with NVSS (NRAO VLA Sky Survey, \citetads{1998AJ....115.1693C}) postage stamps are also shown for comparison. Many of the chosen sources do not lie within the Extended Groth Strip, and are therefore only imaged at low resolution, but high-resolution follow-up images and observations could be made should the need arise.

@@

\pg
Before launching the EGS imaging run using all calibrated datasets, we measure the impact of direction-dependent effects on known sources in the field. Two direction-dependent effects are expected to dominate: decorrelation and differential gains. The first is modeled by the algorithms used in this thesis, whereas the second is a function of the gain variability as a function of distance from the calibrator. Eight calibrator sources, lying at various distances and directions from the calibrator source, are examined. They are summarised in \cref{table.LOBOS.sources}, which is reproduced below: note that number 6 is actually 3C295.

\begin{table}[h!]
\begin{tabular}{ccccc}
\# & RA [hms]    & Dec [dms]   & Dist. from EGS [deg.] & Dist. from 3C295 [deg.] \\\hline
1  & 14:30:18.72 & 52:17:29.80 & 2.041                         & 2.904 \\
2  & 14:19:44.44 & 54:23:04.58 & 1.928                         & 2.517 \\ 
3  & 14:21:20.05 & 53:03:46.00 & 0.864                         & 1.743 \\
4  & 14:21:09.41 & 51:22:32.46 & 1.294                         & 1.728 \\
5  & 14:11:50.32 & 52:49:02.66 & 0.844                         & 0.619 \\
6  & 14:11:20.23 & 52:12:04.30 & 0.915                         & 0.000 \\
7  & 14:08:07.00 & 52:55:11.36 & 1.409                         & 0.869 \\
8  & 14:08:09.76 & 52:44:46.56 & 1.354                         & 0.680 \\
\end{tabular}
\caption{\label{table.LOBOS.sources11}Table recapitulating the positions of all 8 chosen calibrator sources, along with their distance from the observation phase centre (EGS phase centre) and calibrator phase centre (3C295), respectively. Number 6 is 3C295.}
\end{table}
\begin{figure}[h!]
\includegraphics[width=\linewidth]{images/{image_full_ampphase_di_m.NS_Smooth.noise01.fitslbcs.egs.positions}.png}
\caption{Position of the calibrator sources around the EGS, overlaid on the widefield EGS image. LBCS6 is 3C295.}
\label{lbcs-coverage-image11}
\end{figure}
%\begin{figure}[h!]
%\includegraphics[width=0.8\linewidth]{images/EGS_LOBOS_scatterplot}
%\caption{Position of the calibrator sources around the EGS. Note that this is not projected properly onto the celestial sphere.}
%\label{bootes-coverage-image}
%\end{figure}

\pg
If directional gains are measured to have little or no impact, then direction-dependent calibration of the international stations will not be required. This is the ideal scenario, as it does not require the use of fringe fitting \citepads{1984ARA&A..22...97P}. If they are measured to have a significant impact, then direction-dependent calibration will be required for international LOFAR. We expect the angular scale of differential gain evolution to be of the order of a degree, and so most of the EGS should be relatively unaffected.
%
%\pg
%Having quantified the impact of directional gain errors on our field, we then proceed to a patchwise imaging of the EGS.

### 2018-07-18 source/Ch2-techniques/RIME.tex
final commit for sake of backup - thesis completed

\pg Smearing increases with baseline length ($\uvec_{pq}$) and distance from phase centre ($l,m$). The noise amplitude, however, does not decrease. This means that smearing results in a decrease in sensitivity, which becomes significant when attempting to do wide-field, high-resolution images (see \cref{section.EGS.hires}).

>>>>

\pg Smearing increases with baseline length ($\uvec_{pq}$) and distance from phase centre ($l,m$). The noise amplitude, however, does not decrease. This means that smearing results in a decrease in sensitivity, which becomes significant when attempting to do wide-field, high-resolution images (see \cref{section.decorr}).

### 2018-07-18 source/Ch2-techniques/calibration.tex
final commit for sake of backup - thesis completed

\section{Calibration Methods in Radio Interferometry}\label{section.calibration}
\pg
In this section, we will discuss the implementation of interferometric array calibration. Our analysis is based on the RIME formalism, described in \cref{section.RIME}. One key metric of calibration quality is the \emph{dynamic range}, mentioned in Sec. \ref{section.clean}. High dynamic ranges mean that a high contrast has been obtained, and fainter sources can be reached. Here, because the difference between thermal noise and artefacts becomes relevant, we redefine dynamic range as follows
\begin{equation}\label{eq.DR}
DR = \frac{flux_{max}}{S_{min},\max(\sigma_{thermal},\sigma_{artefacts})}
\end{equation}
where $S_{max}$ is the flux of the brightest source, $S_{min}$ the flux of the faintest source detectable above the noise, $\sigma_{thermal}$ the thermal noise in the image, and $\sigma_{artefacts}$ the noise associated with calibration artefacts. This gives a definition for dynamic range useful even in fields with a single source. We can thus use it as a metric for calibration quality specifically.

@@

\pg
First-generation calibration methods (1GC methods) consist of open-loop calibration. This relies entirely on instrument stability, and thus imposes significant design constraints on radio telescopes. It consists of briefly observing an external calibrator before and after each observation run to find calibration solutions for those calibrator observations, and then interpolate between them over the ``target" observation. %\footnote{For a concrete example, see \href{http://www.analog.com/en/analog-dialogue/articles/open-loop-calibration-techniques.html}{the open-loop calibration techniques page of analog.com}.}
%\pg
%Phase calibration in the 1GC era `proper' was not necessary, as engineers were capable of ensuring adequate phase stability in contemporary interferometers. Phase was thus calculated relative to a fixed frame of reference, usually the central antenna of a 3-antenna array. 
In RIME terms (see \cref{section.RIME}), this consists of solving for a very basic form of $\Gjones_p$:
\begin{equation}
\Gjones_p = \begin{bmatrix} a_p t + b_p \end{bmatrix} % \begin{bmatrix} e^{2 \pi i \nu ( \phi_p - \phi_0)} \end{bmatrix}
\end{equation}
where $a_p,b_p$ are constants solved for during open-loop calibration and $t$ is time. %, $\nu$ the observing frequency, $\phi$ the phase at an antenna, and $\phi_0$ the phase at the reference antenna.

@@

\pg
Why then not solve for a set of calibration solutions for each source in the calibration model, iteratively adding new sources to this model as they become above the noise as calibration artefacts - introduced by poor calibration - decrease? There are two main reasons for this: firstly, the SNR when calibrating using very faint source models will be poor. This means that gains will likely be extremely noisy, and applying them will result in increased calibration artefacts in the final image. Secondly, even by optimising for signal-to-noise and only using a few of the brightest sources in the field, the problem quickly becomes intractably large in terms of computing power required, provided that standard complex differentiation is used \citepads[see][]{2016ApJS..223....2V}. Wirtinger differentiation \citepads[see][]{2014arXiv1410.8706T} does alleviate this somewhat by allowing the Hessian matrix to be written as a block-diagonal matrix, but is not implemented in all standard direction-dependent calibration software at the time of writing.

>>>>

\section{Calibration Methods in Radio Interferometry}\label{section.calibration}
\pg
In this section, we will discuss the implementation of interferometric array calibration. Our analysis is based on the RIME formalism, described in \cref{section.RIME}. One key metric of calibration quality is the \emph{dynamic range}, mentioned in Sec. \ref{section.clean}. High dynamic ranges mean that a high contrast has been obtained, and fainter sources can be reached. Here, because the difference between thermal noise and artefacts becomes relevant, we redefine dynamic range as follows
\begin{equation}\label{eq.DR}
DR = \frac{S_{max}}{S_{min},\max(\sigma_{thermal},\sigma_{artefacts})}
\end{equation}
where $S_{max}$ is the flux of the brightest source, $S_{min}$ the flux of the faintest source detectable above the noise, $\sigma_{thermal}$ the thermal noise in the image, and $\sigma_{artefacts}$ the noise associated with calibration artefacts. This gives a definition for dynamic range useful even in fields with a single source. We can thus use it as a metric for calibration quality specifically.

@@

\pg
First-generation calibration methods (1GC methods) consist of open-loop calibration. This relies entirely on instrument stability, and thus imposes significant design constraints on radio telescopes. It consists of briefly observing an external calibrator before and after each observation run to find calibration solutions for those calibrator observations, and then interpolate between them over the ``target" observation. %\footnote{For a concrete example, see \href{http://www.analog.com/en/analog-dialogue/articles/open-loop-calibration-techniques.html}{the open-loop calibration techniques page of analog.com}.}
%\pg
%Phase calibration in the 1GC era `proper' was not necessary, as engineers were capable of ensuring adequate phase stability in contemporary interferometers. Phase was thus calculated relative to a fixed frame of reference, usually the central antenna of a 3-antenna array. 
In RIME terms (see \cref{section.RIME}), this consists of solving for a very basic form of $\Gjones_p$:
\begin{equation}
\Gjones_p = a_p \bm{\mathrm{I}} % \begin{bmatrix} e^{2 \pi i \nu ( \phi_p - \phi_0)} \end{bmatrix}
\end{equation}
where $a_p$ is a complex constant solved for during open-loop calibration. By interpolating between the two calibrator observations, it is possible to have a linear time-dependent gain estimate, though it is not much more complex than the above. %, $\nu$ the observing frequency, $\phi$ the phase at an antenna, and $\phi_0$ the phase at the reference antenna.

@@

\pg
Why then not solve for a set of calibration solutions for each source in the calibration model, iteratively adding new sources to this model as they become above the noise as calibration artefacts - introduced by poor calibration - decrease? There are two main reasons for this: firstly, the SNR when calibrating using very faint source models will be poor. This means that gains will likely be extremely noisy, and applying them will result in increased calibration artefacts in the final image. Secondly, even by optimising for signal-to-noise and only using a few of the brightest sources in the field, the problem quickly becomes intractably large in terms of computing power required, provided that standard complex differentiation is used \citepads[see][]{2016ApJS..223....2V}. Wirtinger differentiation \citepads[see][]{2014arXiv1410.8706T,2015MNRAS.449.2668S} does alleviate this somewhat by allowing the Hessian matrix to be written as a block-diagonal matrix, but is not implemented in all standard direction-dependent calibration software at the time of writing.
